An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-11-16 thesis/introduction-and-background.tex
add notes on cyclic reduction

\subsubsection{Shared memory}

@@

Traditionally,
the Thomas algorithm is employed for solving tridiagonal systems.
The algorithm is derived
from the more general Gaussian (LU) algorithm
applied to tridiagonal matrices.
Implementations of this algorithm are relatively straightforward
and compact ~\cite{numericalrecipes}.
Only the non-zero diagonals of the tridiagonal system
are stored,
and the algorithm performs $2N$ steps
to solve the system.
It is also stable for diagonally dominant matrices.
The Thomas algorithm has an algorithmic complexity of $O(N)$,
and is the fastest serial solver for tridiagonal systems.
It is thus well suited to solution by a single CPU thread.
While the algorithm itself is sequential,
in a multithreaded environment,
the different CPU threads may be used to solve
the different independent tridiagonal systems simultaneously.

@@

Two other popular algorithms for solving
tridiagonal systems are the
cyclic reduction (CR) and parallel cyclic reduction (PCR)
algorithms.

\subsection{Other algorithms}

>>>>

\subsubsection{Shared memory}
\label{subsubsec:shared-memory}

@@

\begin{equation} \label{eqn:general-tridiagonal-system}
\begin{pmatrix}
     b_1 & c_1  \\
     a_2 & b_2  &  c_2  \\
         & a_3  &  b_3 &  c_3  \\
         &      &  a_4 &  b_4 &  c_4  \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\begin{bmatrix}
    x_1 \\
    x_2 \\
    x_3 \\
    x_4 \\
    \vdots \\
    \vdots \\
    x_n
 \end{bmatrix}
=
\begin{bmatrix}
   d_1 \\
   d_2 \\
   d_3 \\
   d_4 \\
   \vdots \\
   \vdots \\
   d_{n}
\end{bmatrix}
\end{equation}

Traditionally,
the Thomas algorithm is employed for solving
general linear systems of the form $A\bm{x} = \bm{d}$,
where $A$ is a tridiagonal matrix with diagonals
$\bm{a}$, $\bm{b}$ and $\bm{c}$
[Eq. (\ref{eqn:general-tridiagonal-system})].
The algorithm is derived
from the more general Gaussian (LU) algorithm
applied to tridiagonal matrices.
Implementations of this algorithm are relatively straightforward
and compact ~\cite{numericalrecipes}.
Only the non-zero diagonals of the tridiagonal matrix
are stored,
and the algorithm performs $2N$ steps
to solve the system.
It is also stable for diagonally dominant matrices.
The Thomas algorithm has an algorithmic complexity of $O(N)$,
and is the fastest serial solver for tridiagonal systems.
It is thus well suited to solution by a single CPU thread.
While the algorithm itself is sequential,
in a multithreaded environment,
the different CPU threads may be used to solve
the different independent tridiagonal systems simultaneously.

@@

Two other popular algorithms for solving
tridiagonal systems are the
cyclic reduction (CR) and
the related parallel cyclic reduction (PCR)
algorithms.

\begin{figure}
\begin{center}
\includegraphics[height=250pt]{img/cyclic-reduction.eps}
\end{center}
\caption{Cyclic reduction.}
\label{fig:cyclic-reduction}
\end{figure}

The cyclic reduction algorithm consists of two phases:
\emph{forward reduction} and \emph{backward substitution}
(Fig. \ref{fig:cyclic-reduction}).
In the forward reduction phase,
every even-indexed equation $i$
is expressed as a
linear combination of equations $i$, $i-1$ and $i+1$,
yielding a new tridiagonal system of
$n/2$ equations in $n/2$ unknowns
[Eqs. (\ref{eqn:forward-reduction-1}) - (\ref{eqn:forward-reduction-4})].
The process is repeated until a system of
2 equations in 2 unknowns is left.

\begin{align} 
& a^{\prime}_i = -a_{i-1}k_1 \
    \label{eqn:forward-reduction-1}& \\
& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 \
    \label{eqn:forward-reduction-2}& \\
& c^{\prime}_i = -c_{i+1}k_2 \
    \label{eqn:forward-reduction-3}& \\
& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 \
    \label{eqn:forward-reduction-4}&
\end{align}
%
where,

\begin{align}
& k_1 = \frac{a_i}{b_{i-1}} \label{eqn:k1-update}& \\
& k_2 = \frac{c_i}{b_{i+1}} \label{eqn:k2-update}&
\end{align}

The 2-by-2 system of equations is solved trivially,
yielding $x_n$ and $x_{n/2}$.
In the backward substitution phase,
every odd-indexed unknown $x_i$ is solved for by
substituting the known values of $x_{i-1}$ and $x_{i+1}$
[Eq. (\ref{eqn:backward-substitution})].

\begin{align} \label{eqn:backward-substitution}
x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - \
    c^{\prime}_ix_{i+1}}{b^{\prime}_i}
\end{align}
%
For the last index $i=n$,
the forward reduction step is instead:
\begin{align} \label{eqn:forward-reduction-last}
    & a^{\prime}_n = -a_{n-1}k_1 & \\
    & b^{\prime}_n = b_n - c_{n-1}k_1 & \\
    & d^{\prime}_n = d_n - d_{n-1}k_1&
\end{align}
%
And for $i=1$, the backward substitution step is instead:

\begin{align} \label{eqn:backward-substitution-first}
x_1 = \frac{d^{\prime}_1 - c^{\prime}_1x_{2}}{b^{\prime}_1}
\end{align}

Thus, in the best case ($n$ parallel processors),
cyclic reduction requires 
$2log_2(n) - 1$ steps.
For even moderately large $n$,
this is significantly smaller than
the $2n$ steps required by the Thomas algorithm.
This makes cyclic reduction a good fit
for massively parallel architectures like GPUs.

\subsection{Parallel cyclic reduction}

The parallel cyclic reduction (PCR) algorithm ~\cite{Zhang2010FTS}
has only the forward reduction phase.
The first forward reduction step is applied to
the odd and even indexed equations seperately,
yielding two reduced systems
of size $n/2$.
Forward reduction applied to both of these
systems then yields \emph{four} reduced
systems of size $n/4$.
The process is repeated till
$n/2$ 2-by-2 systems are left,
all of which can be solved trivially.
The PCR algorithm requires half the number of steps
required by CR ($log_2(n)$),
but does significantly more computation per-step.
Further, unlike CR,
PCR can be implemented free of bank conflicts.

\subsection{Cyclic reduction implementation on GPUs}

The algorithm proposed in this work is based on cyclic reduction,
so it is pertinent to discuss the
implementation of cyclic reduction on GPUs,
the associated issues,
and the relevant literature.

\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/gpu-mapping.eps}
\end{center}
\caption{Mapping work to blocks and threads:
systems are mapped to blocks and
indices are mapped to individual threads.}
\label{fig:gpu-mapping}
\end{figure}

In the GPU implementation of cyclic reduction,
blocks are assigned to tridiagonal systems
(when solving multiple systems),
and threads within a block 
are assigned to equations, or \emph{indices}
(Fig. \ref{fig:gpu-mapping}).
In this way, several grid lines are solved concurrently
by the GPU.
During the forward reduction phase,

\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/forward-reduction-step.eps}
\end{center}
\caption{Updating $\bm{b}$ in the first forward reduction step.}
\label{fig:forward-reduction-step}
\end{figure}

the threads assigned to each even index $i$
compute the coefficients
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$.
In practice, the coefficients are updated \emph{in-place}.
Figure \ref{fig:forward-reduction-step} shows the updates
to the coefficient array $\bm{b}$
in the first forward reduction step).
In each step of forward reduction,
a thread accesses values from the coefficient arrays
$\bm{a}$, $\bm{b}$ and $\bm{c}$ in a \emph{strided} fashion.
At every subsequent step,
this stride is doubled, while the number of active threads is halved
\ref{fig:cyclic-reduction}.
In the backward substitution phase,
the strides are halved at each step,
while the number of active threads is doubled.

This \emph{hourglass} pattern results in low GPU utilization
towards the end of forward reduction
and the beginning of back substitution.
Further,
the long-strided memory accesses towards the end
of the forward reduction
and the beginning of backward substitution
is problematic for both
global memory and shared memory implementations.
The global memory access is \emph{uncoalesced},
leading to severe performance penalties.
The use of shared memory can prevent
long-strided global memory access.
Unfortunately,
the power-of-two strides at each successive
leads to bank conflicts,
as described in Sec. \ref{subsubsec:shared-memory}.
The bank conflicts become increasingly severe
at each forward reduction phase,
and decreasingly so during the back substitution phase.
The limited amount of shared memory places restricions
on the size of tridiagonal systems that can be solved,
and also on the number of blocks that can run in parallel.

Despite these problems,
cyclic reduction remains an attractive algorithm
for GPUs,
for its low algorithmic complexity,
and lower work per step compared to PCR.
Much work has been done on
optimizing cyclic reduction performance on GPUs,
and addressing the issues discussed.
Zhang et al. \cite{Zhang2010FTS} propose a
\emph{hybrid} solver
that uses both cyclic reduction and parallel cyclic reduction
to reduce bank conflicts.
G{\"o}dekke et al. \cite{GoSt11CR}
use a method of separately storing
the even and odd indexed equations
to arrive at a bank-conflict free solver
at the cost of additional shared memory usage.
Davidson et al. \cite{davidson2011register}
describe the method of
\emph{register packing}---performing more computations
on \emph{registers}, rather than shared memory---as
a means to reduce shared memory usage in cyclic reduction.
Esfahanian et. al, \cite{esfahanian2014efficient}
avoid shared memory (and associated bank conflicts entirely)
using a data rearrangement scheme to improve global memory access.
Our approach takes a different route,
and is focussed on near-Toeplitz systems---exploiting
the relatively simple matrix structure
to reduce the number of computations and memory accesses
performed by each thread
at every cyclic reduction step.
